Phylogenetic study of Asgard Archaea

### 2020-07-30 MSc_Dissertation_B156476.tex
Added explanation of NMF, ICA and PCA; diagram from Sompairac

\section{Introduction}
\emph{Say something about ovarian cancer!   An image perhaps.}

\subsection{Matrix factorization methods}
\label{sec-matrix-factorization-intro}

@@

There are three matrix factorization methods of interest to this project:
\begin{enumerate}
\item Independent component analysis (ICA)
\item Non-negative matrix factorization (NMF)
\item Principal component analysis (PCA)
\end{enumerate}

@@

The number of factors, $K$, to extract is a key decision in any matrix factorization approach.  In this regard, PCA differs from ICA and NMF.  In PCA it is reasonable to extract all factors, setting $K=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.   That approach is not reasonable for ICA and NMF, since differing sets of factors will be obtained for different $K$, and for a given choice of $K$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.

@@

\subsection{Ovarian cancer sub-type prediction}

@@

Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $W H$. The following parameters were explored in each case:
\begin{description}
\item[NMF:] Parameters {\tt max\_iter} (algorithm iterations) and {\tt tol} (tolerance of convergence) were optimised for good accuracy and acceptable execution time.  Other parameters of interest are {\tt alpha} (multipluer for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).  L1 regularization  favours elements being precisely zero, whereas L2 regularization will encourage them to be small.  These parameters have not been explored in this project, and the default {\tt alpha = 0} (no regularization) was used.
\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above.  Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
\item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.  However, {\tt sklearn.decomposition.PCA} uses a more efficient 'randomized' algorithm when the given matrix is larger than 500 in both dimensions.  Thus in our use case PCA is seen to have stochastic behaviour.

\end{description}

\subsection{Metagene selection by cluster coherence}
\label{sec-metagene-selection}
As outlined in the discussion, deciding on the number of components (metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

>>>>

\section{Introduction}
High grade serous ovarian cancer (HGSOC) effects the epithelium of the ovaries and fallopian tube.   ``Serous" refers to the epithelial membrane which secretes serous (serum) fluid.  It is the 7th most common cancer in women worldwide .  Epithelial ovarian cancers are classified into two broad subtypes:  Type I are low grade with generally good prognosis; type II are more aggressive and typically carry P53 mutations with defects in mechanisms of DNA repair.  Type II is more-or-less synonomous with HGSOC. \cite{Lisio2019}.

This work is concerned with analysing the patters gene expression in HGSOC dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.

\subsection{Matrix factorization}
\label{sec-matrix-factorization-intro}

@@

The three matrix factorization methods of interest to this project are listed below.  All are forms of dimensionality reduction, but have differing criteria they aim to optimise.  The number of target dimensions is referred to as the rank, $k$.
\begin{description}
\item [Non-negative matrix factorization (NMF).]  
As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which themselves +ve or zero.   This property make the resulting factorization more interpretable since the components are strictly additive.   It also allows for input samples to be directly assigned to one of $k$ clusters.   A good description of NMF and its applications in bioinformatics can be found in \cite{Brunet2004}.
%
\item [Independent component analysis (ICA).] The goal of ICA is to represent the given matrix with components which are statistically independent of each other.  It was originally proposed to solve the blind source separation problem \cite{Comon1994}. If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.    Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.  See \cite{Sompairac2019} for a full explanation, from which figure \ref{fig-pca-ica-nmp-explanation} is taken.
\item [Principal component analysis (PCA).]  PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.  It is the ideal, efficient solution for multivariate Gaussian distributed data.
\end{description}

The contrasting results of the three methods is illustrated in figure.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4, draft=false]{Sompairac_ICA_NMF_PCA.png}
\end{center}
\caption{Illustrating the differing results of component extraction by PCA, ICA and NMF.  From \cite{Sompairac2019}}
\label{fig-pca-ica-nmp-explanation}
\end{figure}

@@

The number of factors, or rank $K$, to extract is a key decision in any matrix factorization approach.  In this regard, PCA differs from ICA and NMF.  In PCA it is reasonable to extract all factors, setting $K=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.   However, that approach is not valid for ICA and NMF, since differing sets of factors will be obtained for different $K$, and for a given choice of $K$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.

@@

@@

Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $W H$. The following parameters were explored in each case:
\begin{description}
\item[NMF:] Parameters {\tt max\_iter} (algorithm iterations) and {\tt tol} (tolerance of convergence) were optimised for good accuracy and acceptable execution time.  Other parameters of interest are {\tt alpha} (multipluer for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).  L1 regularization  favours elements being precisely zero, whereas L2 regularization will encourage them to be small.  These parameters have not been explored in this project, and the default {\tt alpha = 0} (no regularization) was used.
\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above.  Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
\item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.  However, {\tt sklearn.decomposition.PCA} uses a more efficient 'randomized' algorithm when the given matrix is larger than 500 in both dimensions.  Thus in our use case PCA is seen to have stochastic behaviour.
\end{description}

\subsection{Metagene selection by cluster coherence}
\label{sec-metagene-selection}
As outlined in the discussion, deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ vectors. These would appear as separate cluster in repeated sampling, and confound attempts to collect and aggregate.  The solution adopted here is to normalise each factor by requiring that the most extreme value -- i.e. having the greatest absolute value -- is positive, the whole factor being negated if this is not the case.  This is arguably over simplistic, but seems to be effective in practice.
